Categories of Optics

### 2018-06-22 paper.tex
Work on filling in the teleological section

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}
\newcommand{\fover}{\textsc{Over}}

@@

\subsection{Teleological Categories}
\label{teleological-categories}

@@

\begin{definition}[Compare {\cite[Definition 5.1]{CoherenceForLenses}}]
  A \emph{teleological category is} a symmetric monoidal category $(\C, \otimes, I)$, equipped with:
  \begin{itemize}
  \item A wide symmetric monoidal subcategory $\C_d$ of \emph{dualisable morphisms}, with an involutive strong symmetric monoidal functor ${(-)}^* : \C_d \to \overline{\C_d^\op}$, where---not finding a standard symbol for such a thing---we mean $\overline{\C_d^\op}$ to be the category with both the direction of the arrows \emph{and} the order of the tensor flipped: ${(A \otimes B)}^* = B^* \otimes A^*$. \todo{Do we want this to be strict?}
  \item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\C_d$.
  \end{itemize}
\end{definition}
Unpacking the definition, $\varepsilon$ being a symmetric monoidal extranatural transformation amounts to the following diagrams in $\C$ commuting:
\[
  \begin{tikzcd}
    X \otimes Y^* \ar[r, "f \otimes Y^*"]  \ar[d, "X \otimes f^*", swap] & Y \otimes Y^* \ar[d, "\varepsilon_Y"] \\
    X \otimes X^* \ar[r, "\varepsilon_X", swap] & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    X^* \otimes X \ar[r, "\sigma"]  \ar[dr, "\varepsilon_{X^*}", swap] & X \otimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    X \otimes Y \otimes Y^* \otimes X^* \ar[r, "X \otimes \varepsilon_Y \otimes X^*"]  \ar[dr, "\varepsilon_{X \otimes Y}", swap] & X \otimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd}
\]
where $f : X \to Y$ is in $\C_d$.

@@

\begin{example}
  \todo{The funny graph example from the coherence paper?}
\end{example}

\begin{definition}
  A \emph{teleological functor} $F : \C \to \D$ is a symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, and such that $F(\varepsilon_X) = \varepsilon_{FX}$ \todo{up to the structure isomorphisms of $F$}
\end{definition}

@@

@@

  \todo{This works fine when $\C$ is not strict}
\end{proof}

@@

\begin{theorem}
  \label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\otimes \to \T$, unique up to isomorphism, such that $Kj \cong F$.
\end{theorem}
\begin{proof}
  We begin by constructing $K$. On objects, define $K(S, S') = FS \otimes (FS')^*$. Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{M,A'})^*)
  \end{align*}
  Which suggests defining
  \begin{align*}
    K\rep{l}{r} = (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}l) \otimes (F(rs_{A', M}))^* )
  \end{align*}
  The diagram for $K\rep{l}{r}$ in $\T$ is as follows:
  \begin{center}
    \input{diagrams/k-generic-optic.tikz}
  \end{center}

  We must show that this is well defined. Suppose we have two optics related by the coend relation:
  \begin{align*}
    \rep{(f \otimes A) l}{r} = \rep{l}{r (f \otimes A')}
  \end{align*}
  Then well definedness is shown by the equivalence of the diagrams
  \begin{center}
    \input{diagrams/k-well-defined-left.tikz}
    \qquad \raisebox{1.5cm}{$=$} \qquad
    \input{diagrams/k-well-defined-right.tikz}
  \end{center}
  using naturality of the symmetry and extranaturality of the counit.
%\todo{maybe do as string diagram}
%  \begin{align*}
%    K\rep{(f \otimes A) l}{r}
%    &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}(f \otimes A)l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F((A \otimes f)s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    % &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(FA \otimes Ff \otimes FM \otimes FA')(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes (\varepsilon_{FM} (Ff \otimes FM)) \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes (\varepsilon_{FN} (FN \otimes (Ff)^*)) \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes \varepsilon_{FN} \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}(A' \otimes f)))^* ) \\
%    &= (FA \otimes \varepsilon_{FN} \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(r(f \otimes A')s_{A',N}))^* ) \\
%    &= K\rep{ l}{r (f \otimes A')}
%  \end{align*}
  \todo{todo: functoriality, monoidality, preserves dualisable morphisms, preserves counit}

  \todo{uniqueness} 
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $MA$.
\end{definition}

Given such an action, we define
\begin{align*}
  \Optic_\M((S, S'), (A, A')) := \int^{M \in \C} \C(S, MA) \times \C(MA', S')
\end{align*}

@@

\begin{proposition}
  If $F : \C \to \D$ is a functor that commutes with the monoidal actions on $\C$ and $\D$, then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
  If $p = \rep{l}{r}$ is lawful, then verifying the first equation is easy:
  \begin{align*}
  \once(\Optic(F)(\rep{l}{r})) 
  &= \once(\rep{(Fl);\phi^{-1}_{M,A} }{\phi_{M,A'};(Fr) }) \\
  &= (Fl);\phi^{-1}_{M,A};\phi_{M,A'};(Fr)\\
  &= (Fl);(Fr)\\
  &= \id_{FS}
  \end{align*}
  where $\phi_{M,A'} : (FM)(FA) \to F(MA)$ is the structure map for $F$ commuting with the actions.

@@

It is worth emphasising that once a monoidal action has been chosen and a concrete description of the corresponding optics found, no further work is needed to show that the result forms a category of optics, with a subcategory of lawful optics. This is especially useful when devising new optic variants, as we do later.

@@

\todo{There may be something generalisable here with ``monoidal factorisation systems''}

@@

\begin{remark}
\todo{todo?} A careful analysis would be needed to describe traversals in some other category, in particular, the description of $UR^*$ does not make sense if $\C$ is not locally cartesian closed.
\end{remark}

\begin{remark}
Traversable functors can be described as a particular class of polynomial functors known as finitary containers. There may be something generalisable here with other classes of polynomial functors.
\end{remark}

@@

\subsection{``Achromatic'' Lenses}
\todo{http://oleg.fi/gists/posts/2018-03-28-achromatic-lens.html}

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
  \Optic((S, S'), (A, A'))
  &= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
\end{align*}

\begin{remark}
  Equivalently, we could let $\M$ be the category of pointed objects of $\C$ and pointed maps between them.
\end{remark}

\todo{Worth writing out?}

\subsection{Affine Traversals}

A folklore variant of a prism is the \emph{affine traversal}, which allows access to a target that may or may not be present.

Suppose $\C$ is cartesian closed, and has binary coproducts. Then let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
\todo{check $\mathsf{Aff}$ is actually a monoidal category}

Define a monoidal action $\mathsf{Aff} \to [\C, \C]$ where $(P, Q)$ acts on $A$ by
\begin{align*}
  (P, Q) \cdot A = P \sqcup (Q \times A)
\end{align*}
so that as expected,
\begin{align*}
  (P', Q') \cdot (P, Q) \cdot A
  &= (P', Q') \cdot  (P \sqcup (Q \times A)) \\
  &= P' \sqcup (Q' \times (P \sqcup (Q \times A))) \\
  &\cong P' \sqcup (Q' \times P) \sqcup (Q' \times Q \times A) \\
  &= (P' \sqcup (Q' \times P) , Q' \times Q) \cdot A \\
  &= ((P', Q') \otimes (P, Q)) \cdot A
\end{align*}

\begin{remark}
  It is important here that the morphisms in $\mathsf{Aff}$ are only those that arise from pairs of morphisms $P \to P'$ and $Q \to Q'$, although in principle there may be other natural transformations between the corresponding functors $(P, Q) \cdot -$ and $(P', Q') \cdot -$.
\end{remark}

\todo{This feels similar to taking some sort of `compositum' of the two actions $\times$ and $\sqcup$, both embed in this category. Marco suggests taking the pushout of the projections into the pullback, calculated in the 2-cat of monoidal categories.}

Now the set of optics $(S, S') \hto (A, A')$ is:
\begin{align*}
  \Optic_{\mathsf{Aff}}((S, S'), (A, A'))
  &= \int^{M \in \mathsf{Aff}} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
  &\cong \int^{P,Q \in \C} \C(S, (P,Q) \cdot A) \times \C((P,Q) \cdot A', S') \\
  &= \int^{P,Q \in \C} \C(S, P \sqcup (Q \times A)) \times \C(P \sqcup (Q \times A'), S') \\
  &\cong \int^{P,Q \in \C} \C(S, P \sqcup (Q \times A)) \times \C(P,S') \times \C(Q \times A', S') \\
  &\cong \int^{Q \in \C} \C(S, S' \sqcup (Q \times A)) \times \C(Q \times A', S') \\
  &\cong \int^{Q \in \C} \C(S, S' \sqcup (Q \times A)) \times \C(Q, \homC(A', S')) \\
  &\cong \C(S, S' \sqcup (\homC(A', S') \times A))
\end{align*}

We note that this contradicts the folklore description of affine traversals as pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$, see~\cite{AffineTraversalPost}. That set is in fact ``too large'': the provided map $S \times A' \to S$ must be defined even on elements of $S$ whose image under the first map $S \to A \sqcup S'$ lies in the right-hand component, which is not the case in our description.

\todo{Do laws? For \emph{lawful} affine traversals, there may not be a difference}

\todo{Give an example of two elements of the set that correspond to the same affine traversal}

@@

Our goal is to show that $A \mapsto A \otimes -$ is a monoidal functor $a : \C \to [\C_T, \C_T]$. A morphism $f : A \to B$ in $\C$ induces a natural transformation $A \otimes - \to B \otimes -$ with components $A \otimes X \to T(B \otimes X)$ given by composing $A \otimes X \to B \otimes X$ with the unit of the monad. This transformation is natural by \todo{todo}.

@@

\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A', S') \\
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \times A', T S') \\
&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A', T S')) \\
&\cong \C_T(S, \homC(A', T S') \monact A) \\
&\cong \C(S, T(\homC(A', T S') \times A))
\end{align*}

Concrete effectful lenses therefore consist of a single morphism $\munzip : S \to T(\homC(A', T S') \times A)$.

\begin{proposition}
\todo{this is sketchy}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((T\munzip \circ -) \otimes A)\munzip 
  \end{align*}
\end{proposition}
  Or, if you prefer do-notation:
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
== return s

@@

We can say more once we make a specific choice of strong monad.

@@

We can find a concrete description of optics for the action $\monact : \C \times \C_{T_Q} \to \C_{T_Q}$:
\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A', S') \\
&= \int^{M \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M \times A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M) \times \C(S \times Q, A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \C(S \times Q, A \times Q) \times \C_{T_Q}((S \times Q) \monact A', S') \\
&\cong \C_{T_Q}(S, A) \times \C_{T_Q}(S \times Q \times A', S')
\end{align*}
By analogy with ordinary lenses, let us call these maps $\mget$ and $\mput$.

@@

>>>>

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fcreate}{\textsc{Create}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}
\newcommand{\fover}{\textsc{Over}}

@@

\subsection{Teleological Categories}\label{sec:teleological-categories}
\todo{I think there are strictness issues in this section, but the idea is correct}

@@

\begin{definition}[Compare {\cite[Definition 5.1]{CoherenceForLenses}}]
  A \emph{teleological category is} a symmetric monoidal category $(\C, \otimes, I)$, equipped with:
  \begin{itemize}
  \item A symmetric monoidal subcategory $\C_d$ of \emph{dualisable morphisms} containing all the objects of $\C$, with an involutive symmetric monoidal functor ${(-)}^* : \C_d \to \overline{\C_d^\op}$, where---not finding a standard symbol for such a thing---we mean $\overline{\C_d^\op}$ to be the category with both the direction of the arrows \emph{and} the order of the tensor flipped: ${(A \otimes B)}^* \cong B^* \otimes A^*$. Note that there is therefore a canonical isomorphism $\phi : I \cong I^*$
  \item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\C_d$.
  \end{itemize}
\end{definition}
Unpacking the definition, $\varepsilon$ being a symmetric monoidal extranatural transformation amounts to the following diagrams in $\C$ commuting:
\[
  \begin{tikzcd}
    X \otimes Y^* \ar[r, "f \otimes Y^*"]  \ar[d, "X \otimes f^*", swap] & Y \otimes Y^* \ar[d, "\varepsilon_Y"] \\
    X \otimes X^* \ar[r, "\varepsilon_X", swap] & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    X^* \otimes X \ar[r, "\sigma"]  \ar[dr, "\varepsilon_{X^*}", swap] & X \otimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd}\]
\[
  \begin{tikzcd}
    X \otimes Y \otimes Y^* \otimes X^* \ar[r, "X \otimes \varepsilon_Y \otimes X^*"]  \ar[dr, "\varepsilon_{X \otimes Y}", swap] & X \otimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    I \otimes I^* \ar[r,"I \otimes \phi"] \ar[dr, swap, "\varepsilon_I"] & I \ar[d, equal] \\
    & I
  \end{tikzcd}
\]
where $f : X \to Y$ is in $\C_d$.

@@

%\begin{example}
%  \todo{The funny graph example from the coherence paper?}
%\end{example}

%There are few surprises in the following definitions.

\begin{definition}
  A \emph{teleological functor} $F : \C \to \D$ is a symmetric monoidal functor that restricts to a functor $F_d : \C_d \to \D_d$ on the dualisable subcategories, commutes with the duality and such that the counits are preserved:
  \[
   \begin{tikzcd}
    F(X \otimes X^*) \ar[r, "\cong"]  \ar[d, "F\varepsilon_X", swap] & FX \otimes FX^* \ar[d, "\varepsilon_{FX}"] \\
    FI \ar[r, "\cong", swap] & I
  \end{tikzcd}
  \]
\end{definition}

%\begin{definition}
%  A \emph{teleological natural transformation} $\alpha : F \Rightarrow G$ is a monoidal natural transformation that is additionally compatible with the dualisation:
%  \[
%  \begin{tikzcd}
%    (FX)^* \ar[r, "\cong"]  \ar[d, "(\alpha_X)^*", swap] & F(X^*) \ar[d, "\alpha_{X^*}"] \\
%    (GX)^* \ar[r, "\cong", swap] & G(X^*)
%  \end{tikzcd}
%  \]
%  whenever $\alpha_X$ is dualisable. \todo{Does adding this condition make sense?}
%\end{definition}

@@

The definition of teleological category suggests a string diagram calculus similar to that for compact closed categories, but where only counits exist, and only morphisms known to be dualisable may be passed around a counit by extranaturality. We have of course not proven that such a calculus is sound for teleological categories, but we trust that a sceptical reader could verify our arguments equationally.

@@

%  \todo{This works fine when $\C$ is not strict}
\end{proof}

@@

\begin{theorem}
  \label{optic-is-free-teleological-cat}
  Suppose $\C$ is a strict symmetric monoidal category and $\T$ is a teleological category. For all symmetric monoidal functors $F : \C \to \T_d$, there exists a teleological functor $K : \Optic_\otimes \to \T$, unique up to teleological natural isomorphism with the property $Kj \cong F$.
\end{theorem}
\begin{proof}
  We begin by constructing $K$. Note that any $(S, S')$ in $\Optic_\otimes$ can be written as $j(S) \switched j(S')^*$, so on objects define $K(S, S') = FS \otimes (FS')^*$. Suppose $\rep{l}{r} : (S, S') \hto (A, A')$ is an optic. By the previous Proposition,
  \begin{align*}
    \rep{l}{r} = ((A, I) \switched \varepsilon_{(M, I)} \switched (I, A'))(j(s_{M,A}l) \switched j(rs_{A', M})^*)
  \end{align*}
  So up to isomorphism we are forced to define
  \begin{align*}
    K\rep{l}{r} = (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}l) \otimes (F(rs_{A', M}))^* )
  \end{align*}
  The diagram for $K\rep{l}{r}$ in $\T$ is as follows:
  \begin{center}
    \input{diagrams/k-generic-optic.tikz}
  \end{center}

%  \begin{align*}
%    K\rep{(f \otimes A) l}{r}
%    &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F(s_{M,A}(f \otimes A)l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(F((A \otimes f)s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    % &= (FA \otimes \varepsilon_{FM} \otimes (FA')^*)(FA \otimes Ff \otimes FM \otimes FA')(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes (\varepsilon_{FM} (Ff \otimes FM)) \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes (\varepsilon_{FN} (FN \otimes (Ff)^*)) \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}))^* ) \\
%    &= (FA \otimes \varepsilon_{FN} \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(rs_{A',M}(A' \otimes f)))^* ) \\
%    &= (FA \otimes \varepsilon_{FN} \otimes (FA')^*)(F(s_{N,A}l) \otimes (F(r(f \otimes A')s_{A',N}))^* ) \\
%    &= K\rep{ l}{r (f \otimes A')}
%  \end{align*}

There are now a lot of things to check. 
\begin{itemize}
\item Well-definedness: Suppose we have two optics related by the coend relation:
  \begin{align*}
    \rep{(f \otimes A) l}{r} = \rep{l}{r (f \otimes A')}
  \end{align*}
  Then well-definedness is shown by the equivalence of diagrams
  \begin{center}
    \input{diagrams/k-well-defined-left.tikz}
    \qquad \raisebox{1.5cm}{$=$} \qquad
    \input{diagrams/k-well-defined-right.tikz}
  \end{center}
  using naturality of the symmetry and extranaturality of the counit.
\item Functoriality: We have an equivalence of diagrams
  \begin{center}
    \input{diagrams/k-functorial-left.tikz}
    \quad \raisebox{1.5cm}{$=$} \quad
    \input{diagrams/k-functorial-right.tikz}
  \end{center}
using naturality of the symmetry, and monoidality of the counit.
\item Preservation of duals:
\begin{align*}
K((S, S')^*) 
= K(S', S) 
= FS' \otimes (FS)^*
\cong (FS \otimes (FS')^*)^*
= (K(S, S'))^*
\end{align*}
\item Monoidality: We have the isomorphism
\begin{align*}
K((S, S') \switched (T, T')) 
&= K(S \otimes T, T' \otimes S) \\
&= F(S \otimes T) \otimes F(T' \otimes S')^* \\
&\cong FS \otimes FT \otimes (FS')^* \otimes (FT')^* \\
&\cong FS \otimes (FS')^* \otimes FT \otimes (FT')^* \\
&= K(S, S') \otimes K(T, T')
\end{align*}
using the symmetry of $\T$, and
\begin{align*}
K(I, I) 
&= FI \otimes (FI)^* \\
&\cong I \otimes I^* \\
&\cong I
\end{align*}
That these obey the required coherences is straightforward. More difficult is showing that the first isomorphism is natural in $(S, S')$ and $(T, T')$: \todo{todo}
\item Preservation of dualisable morphisms:  For a morphism $\iota(f, g)$:
\begin{align*}
K(\iota(f, g))
    &= K(\rep{\lambda_A^{-1} f}{g \lambda_{A'}}) \\
    &= (FA \otimes \varepsilon_{FI} \otimes (FA')^*)(F(s_{I,A}\lambda_A^{-1} f) \otimes (F(g \lambda_{A'}s_{A', I}))^* ) \\
    &= (FA \otimes (FA')^*)(Ff \otimes (Fg)^* ) \\
    &= Ff \otimes (Fg)^*
\end{align*}
and this is dualisable, as dualisability is preserved by taking the monoidal product and duals.
\item Preservation of counits: 
\begin{align*}
K(\varepsilon_{(S, S')}) 
&= K(c_{S \otimes S'}) \\
&= K(\rep{\rho_{S \otimes S'}^{-1}}{\rho_{S \otimes S'}}) \\
&= (FI \otimes \varepsilon_{F(S \otimes S')} \otimes (FI)^*)(F(s_{S \otimes S',I}\rho_{S \otimes S'}^{-1}) \otimes (F(\rho_{S \otimes S'} s_{I, S \otimes S'}))^* ) \\
&= (\varepsilon_{F(S \otimes S')})(F(S \otimes S') \otimes F(S \otimes S')^* ) \\
&= \varepsilon_{F(S \otimes S')}
\end{align*}
by strictness of the monoidal unit.
\todo{Show this is actually the morphism we want}
\end{itemize}
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.
\end{definition}

Given such an action, we define
\begin{align*}
  \Optic_\M((S, S'), (A, A')) := \int^{M \in \M} \C(S, MA) \times \C(MA', S')
\end{align*}

@@

\begin{proposition}
  If $F : \C \to \D$ is a functor that commutes with the monoidal actions on $\C$ and $\D$, then $\Optic(F) : \Optic_\M \to \Optic_\N$ restricts to a functor $\Lawful_\M \to \Lawful_\N$.
\end{proposition}
\begin{proof}
  If $p = \rep{l}{r}$ is lawful, then verifying the first equation is easy:
  \begin{align*}
  \outside(\Optic(F)(\rep{l}{r})) 
  &= \outside(\rep{(Fl);\phi^{-1}_{M,A} }{\phi_{M,A'};(Fr) }) \\
  &= (Fl);\phi^{-1}_{M,A};\phi_{M,A'};(Fr)\\
  &= (Fl);(Fr)\\
  &= \id_{FS}
  \end{align*}
  where $\phi_{M,A'} : (FM)(FA) \to F(MA)$ is the structure map for $F$ commuting with the actions.

@@

It is worth emphasising that once a monoidal action has been chosen and a concrete description of the corresponding optics found, no further work is needed to show that the result forms a category of optics with a subcategory of lawful optics. This is especially useful when devising new optic variants, as we do later.

@@

%\todo{There may be something generalisable here with ``monoidal factorisation systems''}

@@

\begin{remark}
A careful analysis would be needed to describe traversals in some other category, in particular, the description of $UR^*$ does not make sense if $\C$ is not locally cartesian closed.
\end{remark}

\begin{remark}
Traversable functors can be described as a particular class of polynomial functors known as finitary containers. It may be possible to generalise this to other classes of polynomial functors.
\end{remark}

@@

@@

Our goal is to show that the assignment $A \mapsto A \otimes -$ extends to a monoidal functor $a : \C \to [\C_T, \C_T]$. A morphism $f : A \to B$ in $\C$ induces a natural transformation $A \otimes - \to B \otimes -$ with components $A \otimes X \to T(B \otimes X)$ given by composing $A \otimes X \to B \otimes X$ with the unit of the monad. This transformation is natural by \todo{todo}.

@@

\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A', S') \\
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \otimes A', T S') \\
&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A', T S')) \\
&\cong \C_T(S, \homC(A', T S') \monact A) \\
&\cong \C(S, T(\homC(A', T S') \otimes A))
\end{align*}

Concrete effectful lenses therefore consist of a single morphism $\munzip : S \to T(\homC(A', T S') \otimes A)$. The category $\C$ includes into $\C_T$ in a way that preserves the action of $\C$, so there is an induced inclusion $\Optic_\otimes \to \Optic_\monact$.

\begin{proposition}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\eta_{\homC(A, TS) \otimes A)}\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((\munzip \circ_T -) \otimes A)\munzip 
  \end{align*}
  where \[ \munzip \circ_T - : \homC(A, TS) \to \homC(A, T(\homC(A, TS) \otimes A)) \] denotes internal Kleisli composition and \[\coev_{\homC(A, TS), A} : \homC(A, TS) \to \homC(A, \homC(A, TS) \otimes A) \] is coevaluation.
\end{proposition}
  Or, if you prefer do-notation:
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
== 
return s

@@

If we choose a specific monad, we can hope to simplify the description of concrete effectful lenses and their laws.

@@

Working in a cartesian closed category, we can find a concrete description of optics for the action $\monact : \C \times \C_{T_Q} \to \C_{T_Q}$ that is closer to that for ordinary lenses.
\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A', S') \\
&= \int^{M \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M \times A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M) \times \C(S \times Q, A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \C(S \times Q, A \times Q) \times \C_{T_Q}((S \times Q) \monact A', S') \\
&\cong \C_{T_Q}(S, A) \times \C_{T_Q}(S \times Q \times A', S')
\end{align*}
By analogy with ordinary lenses, let us call these maps $\mget$ and $\mput$.

@@

\subsection{Further Examples}
\todo{I probably don't have the standing to get away with saying} The dedicated reader may enjoy re-deriving the concrete representation and laws for the following optic varieties:

\begin{itemize}
\item \emph{``Achromatic'' Lenses} \cite[Section 5.2]{ProfunctorOpticsThesis} are lenses that also admit an operation $\fcreate : A \to S$. These are optics for the action of $\C$ on itself by $M \cdot A = (M \sqcup 1) \times A$, or equivalently, of the category of pointed objects of $\C$ on $\C$ by cartesian product. Concrete achromatic lenses $S \hto A$ are elements of the set \[\C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S').\]
%\begin{align*}
%  \Optic((S, S'), (A, A'))
%  &= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
%  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
%  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
%  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
%\end{align*}
\item \emph{Affine Traversals} \cite{AffineTraversalPost} allow access to a target that may or may not be present. Suppose $\C$ is cartesian closed and has binary coproducts. Let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
The category $\mathsf{Aff}$ acts on $\C$ by $(P, Q) \cdot A = P \sqcup (Q \times A)$. A concrete representation of the optics for this action is \[\C(S, S' \sqcup (\homC(A', S') \times A)).\]

Affine traversals are described in the folklore as pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$. Such a pair does determine an affine traversal, but in fact gives more information than is necessary: the right-hand map need not be defined at all $S$.
